A Study of Institutional Spending on Open Access Publication Fees in Germany

### 2016-06-10 article.Rmd
proofreading methods section

We analysed self-reported cost data released by the Open APC initiative on May 13, 2016,[^24] to assess institutional spending on open access publication fees in Germany. In addition to administrative data about the amount paid per article including value added tax, the reporting institution, and the year of invoicing, we used information about whether an article was published in a fully open access journal or in a hybrid journal, as well as the DOI reported in the data-set.

We fetched bibliographic metadata for each article from Crossref on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from Crossref as well, this information was retrieved at the time when the participating institutions submitted their expenditure. The Open APC initiative kept track of the date when these data-sets were submitted with Git, a version control system, increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub to be transparent over time. However, during these data collection activities Crossref had regularly updated metadata to represent ongoing mergers of publishing houses. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve  updated bibliographic metadata from Crossref, and to merge these records with the administrative information rather than re-using the historical publisher and journal information contained in the Open APC data-set.

We used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access Crossref's REST API[^25]. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types, the International Standard Serial Number used to identify journals, were distinguished. It also contained normalised publisher information, thus avoiding confusion about naming of publishing houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Because Crossref was not the only registration agency for DOIs, but also the agencies DataCite and Medra minted DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

Data collection also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by Germany universities and research organisations with that of other initiatives, we consulted the openly available data-sets from the the Austrian Science Fund FWF [@fwf_apc_14; @fwf_apc_15], Jisc [@jisc_14; @jisc_15] and the Wellcome Trust [@wellcome_apc_13_14; @wellcome_apc_14_15]. For analysis, we obtained the overall publication fee spending on both fully and hybrid open access journal articles. In the case of FWF, we gathered the cost information from the accompanying spending reports. We used the spreadsheet data to calculate Wellcome Trust's and Jisc's spending, and converted the prices from GBP to Euro in accordance with the average foreign exchange reference rates provided by the European Central Bank. Our comparison of the open data initiatives focussed on the last two years 2014 and 2015. Because Wellcome Trust's spending was reported for the fiscal periods 2013 - 2014 and 2014 - 2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded articles from the analysis with missing information about the cost or the journal type. In the case of Jisc's 2014 data [@jisc_14], for instance, we excluded spending on 2,812 publications that amounted to 4.861.772 € from the analysis because no publication type was given in the data-set.

Data curation activities of the Open APC initiative and those of the other initiatives differed in some respects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institutions were required to report, in case of the Wellcome Trust spending data, DOIs were additionally identified by automated compliance checks. Our first screening of the data-sets revealed that some articles published in Crossref-indexed journals lacked a DOI. Because of these different methods to curate the cost data, and as our main focus was institutional funding for publication fees in Germany, we decided to only compare German spending with that reported by other initiatives. We, therefore, did not analyse the distribution of spending over publishers and journal titles, or the indexing coverage in Crossref for the Austrian and UK spending data.

>>>>

We analysed self-reported cost data released by the Open APC initiative on May 13, 2016,[^24] to assess institutional spending on open access publication fees in Germany. In addition to administrative data about the amount paid per article, including value-added tax, the reporting institution, and the year of invoicing, we used information about whether an article was published in a fully open access journal or in a hybrid journal, as well as the DOI reported in the data-set.

We obtained bibliographic metadata for each article from Crossref on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from Crossref as well, this information was retrieved at the time when the participating institutions submitted their expenditure. To be transparent over time, the Open APC initiative kept track of the date when these data-sets were submitted with Git, a version control system that is increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub  However, during these data collection activities, Crossref regularly updated the metadata to represent ongoing mergers of publishing houses. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, which operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve  updated bibliographic metadata from Crossref and to merge these records with the administrative information rather than re-using the historical publisher and journal information contained in the Open APC data-set.

We used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access Crossref's REST API.[^25] We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types (the International Standard Serial Number used to identify journals) were distinguished. This source also contained normalised publisher information, thus avoiding confusion regarding the naming of publishing houses that other studies faced when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Because Crossref is not the only registration agency for DOIs – the agencies DataCite and Medra also mint DOIs for scholarly works – we also identified the DOI agency for each article with the help of the rcrossref client.

Data collection also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by German universities and research organisations with that of other initiatives, we consulted the openly available data-sets from the Austrian Science Fund (FWF) [@fwf_apc_14; @fwf_apc_15], Jisc [@jisc_14; @jisc_15] and the Wellcome Trust [@wellcome_apc_13_14; @wellcome_apc_14_15]. For analysis, we obtained the overall publication fee spending on both fully and hybrid open access journal articles. In the case of FWF, we gathered the cost information from the accompanying spending reports. We used the spreadsheet data to calculate Wellcome Trust's and Jisc's spending, and converted the prices from GBP to Euro in accordance with the average foreign exchange reference rates provided by the European Central Bank. Our comparison of the open data initiatives focussed on the last two years: 2014 and 2015. Because Wellcome Trust's spending was reported for the fiscal periods 2013-2014 and 2014-2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded articles from the analysis with missing information about the cost or the journal type. In the case of Jisc's 2014 data [@jisc_14], for instance, we excluded spending on 2,812 publications that amounted to € 4.861.772 from the analysis because no publication type was given in the data-set.

Data curation activities of the Open APC initiative and those of the other initiatives differed in some respects. For instance, the DOI was a mandatory element in the Open APC data template that the participating institutions were required to report, whereas in the case of the Wellcome Trust spending data, DOIs were additionally identified by automated compliance checks. Our first screening of the data-sets revealed that some articles published in Crossref-indexed journals lacked a DOI. Because of these different methods to curate the cost data and because our main focus was institutional funding for publication fees in Germany, we decided to compare only German spending with that reported by other initiatives. We did not, therefore, analyse the distribution of spending over publishers and journal titles or the indexing coverage in Crossref for the Austrian and UK spending data.
